---
title: "Problem Set 8"
subtitle: "STAT E-80: Basic Probability Using R"
output:
  pdf_document: default
  html_document:
    df_print: paged
---


# Instructions

**Due:** Wednesday, April 10, 2019

#Instructions

First, save this notebook with a title consisting of "ProblemSet8" and then your last name, with no spaces. So, for Bob Bunny, the file should be called "ProblemSet8Bunny". Also, type your name in the subtitle on line 2 above. Each problem will have a problem statement, followed with space for you to answer. Type your work in the code chunk where it says, "Type your answer in here". If we ask a question that has a numeric answer (for example, "How much money does Bob Bunny owe?"), then you should write some text in the notebook with your answer; don't assume that just the R printout is sufficient.

When you're all done, knit the notebook to a PDF, and upload both the notebook and the PDF to Canvas. (Note: The notebook will have the file extension ".rmd".)

\newpage
# Problem 1

In lecture, we saw the infinite series:
$$
\sum_{k=0}^\infty x^k\ =\ 1 + x + x^2 + \ldots
$$
We used this infinite series to show that the formula for the probability function for the geometric distribution did indeed sum to 1 and thus was a valid probability distribution. In order to derive the expected value of the geometric distribution, we had to use the infinite series:
$$
\sum_{k=0}^\infty k \cdot x^k\ =\ 0 + x + 2x^2 + 3x^3 + \ldots
$$
I stated in the reading that:
$$
\sum_{k=0}^\infty k \cdot x^k\ =\ \frac{x}{(1-x)^2}
$$
Now we're going to explore this formula.


### Part (a)

Set $x = 3/7$. Then use the expression on the right-hand side of the formula to calculate the value of the infinite series.

**Answer**
(3/7)
```{r}
answer=(3/7)/(1-(3/7))^2
cat (answer)
```




### Part (b)

Create a vector of the form:
$$
(0, 1, 2, 3, \ldots, 25)
$$
Then create a vector of the form:
$$
(x^0, x^1, x^2, \ldots, x^{25})
$$
Then use these two vectors to calculate the sum of the first 25 terms of the infinite series:
$$
\sum_{k=0}^{25} k \cdot x^k\ =\ 0 + x + 2x^2 + 3x^3 + \ldots + 25 x^{25}
$$

**Answer**

Vector.1<-seq(1, 25, 1)
```{r}
Vector.1<-seq(1, 25, 1)
print (Vector.1)
vector.squared<-Vector.1^2
print(vector.squared)
```


### Part (c)

Print out a list of statements that display the successive cumulative sums of the terms of the infinite series:

* The first statement should print out the numerical value of 0 x^0, whatever that might be.

* The second statement should print out the numerical value of 0 * x^0 + 1 * x^1, whatever that might be.

* The third statement should print out the numerical value of 0 * x^0 + 1 * x^1 + 2 * x^2, whatever that might be.

I'll get you started:

    The cumulative total of the first 0 terms is: 0 

    The cumulative total of the first 1 terms is: 0.4285714

    The cumulative total of the first 2 terms is: 0.7959184

In the end, you should print out 26 statements.

Hint: you can do this very simply using a `for` loop to accumulate a running total.






**Answer**







```{r}
x<-0.4285714
number.terms<-26
i <- 0

# Now we'll create a vector to keep track
# of all the cumulative totals:

cumulative.total.vector <- numeric( number.terms )

# Now we enter the for loop:

for( n in 0:number.terms ) {
    
    # Each time through the loop, we will add a
    # higher power of x to the total:
    
    i<- (n * x^n)+i
    
    # Now we store the current value of total.sum
    # in the cumulative.total.vector. Notice that
    # we have to add 1 to the power.index:
    
    cumulative.total.vector[ n+1  ] <- i
}

# Now that we're done with the loop, we can 
# display the cumulative.total.vector, which
# will have all the running totals:

cumulative.total.vector
```




\newpage
# Problem 2

In this problem, we will do some simple experiments with the geometric distribution.

Let $X$ denote a random variable that has a geometric distribution with a probability of success $p = 0.2$. 

### Part (a)

Calculate the probability of observing 2 failures before the first success i.e. calculate $\Pr(X = 3)$. Don't use the `dgeom()` function, but instead use the algebraic formula from lecture:
$$
\Pr(X = k)\ =\ (1-p)^k \cdot p
$$

**Answer**

```{r}
answer<-((1-.2)^3)*.2
cat(answer)
```



### Part (b)

Now use the `dgeom()` function to calculate $\Pr(X = 3)$.

**Answer**
```{r}
dgeom(3, 0.2)
```





### Part (c)

Use the formulas from lecture and reading to calculate the expected value and variance of the random variable $X$.

**Answer**

```{r}
answer<-(1-.2)/.2
cat(answer)
```






### Part (d)

Construct a simulation to confirm your calculations in parts (a), (b), and (c):

* Run at least 10,000 simulation replications.

* In each simulation, model the process of generating a geometric random variable by using a `while()` loop to repeatedly sample until the first success. You are welcome to use the code in the Lecture 9 R Notebook.

* Finally, calculate the proportion of simulation where the value of the random variable is 3 (that's parts (a) and (b)), and then report the mean and variance.

**Answer**
```{r}
# Since this is a geometric random variable,
# we'll start by specifying the probability of success:

probability.success <- 0.2

number.replications <- 10000

result.vector <- numeric( number.replications )

for( replication.index in 1:number.replications ) {
    
    # First, we set up the variable that keeps track of
    # the whether or not a success has occurred:
    
    failure.status <- "Failure"
    
    # Now we have an accumulator random variable
    # that will keep track of the number of failures
    # before the first success:
    
    number.failures <- 0
    
    # Now we enter the while loop
    
    while( failure.status == "Failure" ) {
        
        # First, let's run a trial, which will either be
        # a success or a failure:
        
        failure.status <- sample( c("Success", "Failure"), 1,
                               prob = c(probability.success, 1 - probability.success) )
        
        # Now let's check to see if the trial is a failure
        # If the trial is a Failure, then update the 
        # accumulator variable that is counting the number
        # of failures.
        
        if( failure.status == "Failure" ) {
            number.failures <- number.failures + 1
        }
        
    }
    
    result.vector[ replication.index ] <- number.failures
    
}
 answer<-mean( result.vector == 3 )
# cat ("P (X=3)=",answer)
cat("(P (X=3)=",answer, "\n")
varance<-var( result.vector )
 cat ("varance=",varance)
    
```








### Part (e)

Make a histogram of the `result.vector` from your simulation. Make sure you provide a main title, label the x axis, and use the `prob = TRUE` option. Also, choose a cool color from the R Colors document.

**Answer**


```{r}
hist( result.vector,
      main = "Histogram of result vector",
      col = "darkturquoise" )

```






\newpage
# Problem 3

Recall that the *cumulative probability* of a value $k$ is the probability that a random variable is less than or equal to that value; alternatively, we could say that it is the probability that the random variable is at most the value $k$. In the reading, I showed that there is a nice formula for the cumulative probability for a geometric distribution with probability of success $p$:
$$
\Pr(X \leq k)\ =\ 1 - (1-p)^{k+1}
$$

### Part (a)

Let $X$ be a geometric random variable with probability of success $p = 0.06$. Using the formula above, calculate the probability that $X$ is at most 5.

**Answer**
```{r}
1-(1-.06)^6
```





### Part (b)

Use the built-in R function `pgeom()` to calculate the cumulative probability that $X$ is at most 5.

**Answer**


```{r}
pgeom(5, 0.06)
```


### Part (c)

Write a simulation to confirm your calculations in parts (a) and (b). Use a `while` loop and the `sample()` function to explicitly model the process of repeating the process until the first success.

**Answer**Now let's see how to simulate a geometric random variable. Here we will use a `while` loop to repeatedly sample Bernoulli random variables with a probability of success $p = 0.2$:
```{r}
# Since this is a geometric random variable,
# we'll start by specifying the probability of success:

probability.success <- 0.06

number.replications <- 10000

result.vector <- numeric( number.replications )

for( replication.index in 1:number.replications ) {
    
    # First, we set up the variable that keeps track of
    # the whether or not a success has occurred:
    
    failure.status <- "Failure"
    
    # Now we have an accumulator random variable
    # that will keep track of the number of failures
    # before the first success:
    
    number.failures <- 0
    
    # Now we enter the while loop
    
    while( failure.status == "Failure" ) {
        
        # First, let's run a trial, which will either be
        # a success or a failure:
        
        failure.status <- sample( c("Success", "Failure"), 1,
                               prob = c(probability.success, 1 - probability.success) )
        
        # Now let's check to see if the trial is a failure
        # If the trial is a Failure, then update the 
        # accumulator variable that is counting the number
        # of failures.
        
        if( failure.status == "Failure" ) {
            number.failures <- number.failures + 1
        }
        
    }
    
    result.vector[ replication.index ] <- number.failures
}
mean( result.vector <= 5 )

```







### Part (d)

The survival function for a random variable is $\Pr(X > k)$, the probability that a random variable is strictly greater than some value $k$. The formula for the survival function for a geometric random variable is:
$$
\Pr(X > k)\ =\ (1 - p)^{k+1}
$$
Let $X$ be a geometric random variable with probability of success $p = 0.14$. Use this algebraic formula to calculate the probability that $X$ is strictly greater than 8.

**Answer**
```{r}
(1-.14)^9
```



### Part (e)

Use the `pgeom()` function with the `lower.tail == FALSE` option to calculate the probability that a geometric random variable with probability of success $p = 0.14$ is strictly greater than 8.

**Answer**

```{r}
pgeom(8, .14,lower.tail=FALSE)
```



### Part (f)

Construct a simulation to confirm your calculations in parts (d) and (e).

**Answer**
```{r}
# Since this is a geometric random variable,
# we'll start by specifying the probability of success:

probability.success <- 0.14

number.replications <- 10000

result.vector <- numeric( number.replications )

for( replication.index in 1:number.replications ) {
    
    # First, we set up the variable that keeps track of
    # the whether or not a success has occurred:
    
    failure.status <- "Failure"
    
    # Now we have an accumulator random variable
    # that will keep track of the number of failures
    # before the first success:
    
    number.failures <- 0
    
    # Now we enter the while loop
    
    while( failure.status == "Failure" ) {
        
        # First, let's run a trial, which will either be
        # a success or a failure:
        
        failure.status <- sample( c("Success", "Failure"), 1,
                               prob = c(probability.success, 1 - probability.success) )
        
        # Now let's check to see if the trial is a failure
        # If the trial is a Failure, then update the 
        # accumulator variable that is counting the number
        # of failures.
        
        if( failure.status == "Failure" ) {
            number.failures <- number.failures + 1
        }
        
    }
    
    result.vector[ replication.index ] <- number.failures
}
mean( result.vector > 8 )
```





\newpage
# Problem 4

### Part (a)

Suppose $X$ is a geometric random variable, with an unknown probability of success $p$. However, you know that the expected value of the random variable is 7. Determine $p$, the probability of success.

**Answer**
P(X)=(1-p)^k
e(X)=(1-P)/P
7=(1-P)/P
x7=-x-1
8x=1
x=1/8

### Part (b)

Verify your answer in part (a) by generating 10,000 random observations from a geometric distribution with the probability of success that you calculated, and then seeing if the mean of these observations is close to 7. You can use the `rgeom()` function for this.

**Answer**

```{r}
# Since this is a geometric random variable,
# we'll start by specifying the probability of success:

probability.success <-(1/8)

number.replications <- 10000

result.vector <- numeric( number.replications )

for( replication.index in 1:number.replications ) {
    
    # First, we set up the variable that keeps track of
    # the whether or not a success has occurred:
    
    failure.status <- "Failure"
    
    # Now we have an accumulator random variable
    # that will keep track of the number of failures
    # before the first success:
    
    number.failures <- 0
    
    # Now we enter the while loop
    
    while( failure.status == "Failure" ) {
        
        # First, let's run a trial, which will either be
        # a success or a failure:
        
        failure.status <- sample( c("Success", "Failure"), 1,
                               prob = c(probability.success, 1 - probability.success) )
        
        # Now let's check to see if the trial is a failure
        # If the trial is a Failure, then update the 
        # accumulator variable that is counting the number
        # of failures.
        
        if( failure.status == "Failure" ) {
            number.failures <- number.failures + 1
        }
        
    }
    
    result.vector[ replication.index ] <- number.failures
}
expected.value.sim=mean( result.vector )
cat ("simulation gives me ",expected.value.sim )
```

### Part (c)

Suppose $X$ is a geometric random variable, with an unknown probability of success $p$. However, you know that the variance of the random variable is 70/9. Determine $p$, the probability of success.

**Answer**
varience(x)=(1-P)/P^2
70/9=(1-P)/P^2
70/9p^2=(1-p)
70/9p^2-1+p=0
apply quadradic formula
get .3 and -.42
-.42 is stupid
take .3
p=.3

### Part (d)

Verify your answer in part (c) by generating 10,000 random observations from a geometric distribution with the probability of success that you calculated, and then seeing if the variance of these observations is close to 70/9. You can use the `rgeom()` function for this.

**Answer**

```{r}
# Since this is a geometric random variable,
# we'll start by specifying the probability of success:

probability.success <-.3

number.replications <- 10000

result.vector <- numeric( number.replications )

for( replication.index in 1:number.replications ) {
    
    # First, we set up the variable that keeps track of
    # the whether or not a success has occurred:
    
    failure.status <- "Failure"
    
    # Now we have an accumulator random variable
    # that will keep track of the number of failures
    # before the first success:
    
    number.failures <- 0
    
    # Now we enter the while loop
    
    while( failure.status == "Failure" ) {
        
        # First, let's run a trial, which will either be
        # a success or a failure:
        
        failure.status <- sample( c("Success", "Failure"), 1,
                               prob = c(probability.success, 1 - probability.success) )
        
        # Now let's check to see if the trial is a failure
        # If the trial is a Failure, then update the 
        # accumulator variable that is counting the number
        # of failures.
        
        if( failure.status == "Failure" ) {
            number.failures <- number.failures + 1
        }
        
    }
    
    result.vector[ replication.index ] <- number.failures
}
var( result.vector )
```























