---
title: "Problem Set 9 Solutions"
subtitle: 'STAT E-80: Basic Probability Using R'
output:
  pdf_document: default
  html_document:
    df_print: paged
---


# Instructions

**Due:** Wednesday, April 24, 2019

#Instructions

First, save this notebook with a title consisting of "ProblemSet6" and then your last name, with no spaces. So, for Bob Bunny, the file should be called "ProblemSet6Bunny". Also, type your name in the subtitle on line 2 above. Each problem will have a problem statement, followed with space for you to answer. Type your work in the code chunk where it says, "Type your answer in here". If we ask a question that has a numeric answer (for example, "How much money does Bob Bunny owe?"), then you should write some text in the notebook with your answer; don't assume that just the R printout is sufficient.

When you're all done, knit the notebook to a PDF, and upload both the notebook and the PDF to Canvas. (Note: The notebook will have the file extension ".rmd".)







# Problem 1

### Problem Statement

Let $X$ be an exponential random variable with rate parameter $r = 1/8$. Then the density function for $X$ is:
$$
f_X(x)\ =\ \frac{1}{8} \cdot e^{-x/8},\ \ x \geq 0
$$

### Part (a)

Draw a random sample of 10,000 observations from this distribution, and name this sample `problem.1.sample`. Construct a histogram of this sample, and then superimpose the density function using the built-in `curve()` function in R. (Hint: remember to use the `add = TRUE` option.)

**Answer**

```{r}
problem.1.sample <- rexp(10000, rate = 1/8)

hist( problem.1.sample,
      main = "Problem 1 Exponential Sample Histogram",
      xlab = "x",
      ylab = "Density",
      breaks = 50,
      prob = TRUE,
      col = "aquamarine2" )

curve( dexp(x, rate = 1/8),
       lwd = 2,
       add = TRUE)
```


### Part (b)

The cumulative probability function $F_X(x) = \Pr(X \leq x)$ for an exponential random variable with rate parameter $r$ is:
$$
F_X(x)\ =\ \Pr(X \leq x)\ =\ 1 - e^{-r \cdot x},\ \ x \geq 0,\ r \geq 0
$$
Calculate the cumulative probability $F_X(12) = \Pr(X \leq 12)$ for an exponential random variable with rate parameter $r = 1/8$ using this formula. Then do the same calculation using the built-in R function `pexp()`.

**Answer** First, this is the calculation using the algebraic formula:
```{r}
1 - exp( - 1/8 * 12)
```
Now here is the same calculation using the built-in `pexp()` function:
```{r}
pexp(12, rate = 1/8)
```


### Part (c)

Verify your result in part (b) using the simulated data `problem.1.sample` from part (a). (Hint: create a logical vector that is `TRUE` exactly when a value in `problem.1.sample` is less than or equal to 12, and then determine what proportion of the elements in this vector are `TRUE`.)

**Answer**

```{r}
mean( problem.1.sample <= 12 )
```






### Part (d)

The expected value and variance of an exponential random variable with rate parameter $r$ are:
\begin{eqnarray*}
\hbox{E}[X] & = & \frac{1}{r}\\
\\
\hbox{Var}[X] & = & \frac{1}{r^2}
\end{eqnarray*}
Calculate the expected value and variance for the random variable $X$ from part (a).

**Answer**

```{r}
cat( "Expected value:", 1 / (1/8), "\n" )
```

```{r}
cat( "Variance:", 1/(1/8)^2, "\n" )
```



### Part(e)

Use the simulated data stored in `problem.1.sample` to verify your calculations in part (d) of the expected value and variance of $X$.

**Answer**

```{r}
cat( "Expected value:", mean( problem.1.sample ), "\n" )
cat( "Variance:", var( problem.1.sample ), "\n" )
```
The expected value should be close to 8, and the variance should be close to 64.


\newpage
# Problem 2

### Problem Statement

Let $Y$ be a continuous random variable with density function:
$$
f_X(x)\ =\ 3 e^{-3x},\ \ x \geq 0
$$

### Part (a)

What is the name of the probability distribution of $X$? What is the parameter for this distribution?

**Answer** This is an exponential distribution with rate parameter $r = 3$.


### Part (b)

Draw a random sample of 10,000 independent observations from the distribution $Y$, and call this sample `problem.2.sample`. Create a histogram for this distribution, and then superimpose the density curve for the distribution that you named in part (a).

**Answer**

```{r}
problem.2.sample <- rexp( 10000, rate = 3)

hist( problem.2.sample,
      main = "Problem 2 Mystery Distribution Histogram",
      xlab = "X",
      ylab = "Density",
      prob = TRUE,
      breaks = 50,
      col = "salmon1")

curve( dexp(x, rate = 3), lwd = 2, add = TRUE )
```

### Part (c)

The survival function $S_X(s) = \Pr(X > x)$ for an exponential random variable with rate parameter $r$ has the form:
$$
S_X(x)\ =\ \Pr(X > x)\ =\ e^{-r x}
$$
Calculate the survival probability $S_X(0.5) = \Pr(X > 0.5)$ for an exponential random variable with rate parameter $r = 3$ using this formula. Then do the same calculation using the built-in R function `pexp()` (remember the `lower.tail` option!).

**Answer**

```{r}
exp( - 3 * 0.5)
pexp( 0.5, rate = 3, lower.tail = FALSE )
```


### Part (d)

Use your simulated data from part (b) to verify your answer in part (c).

**Answer**

```{r}
mean( problem.2.sample > 0.5 )
```


### Part (e)

Calculate the expected value and variance for the density in part (a).

**Answer** The expected value is:
```{r}
cat( "Expected value:", 1/3, "\n" )
```
The variance is:
```{r}
cat( "Variance:", 1/3^2 )
```



### Part (e)

Use your simulated data from part (a) to verify your calculations of the expected value and variance of $X$ in part (d).

**Answer**

```{r}
cat( "Simulated expected value:", mean( problem.2.sample ), "\n" )
cat( "Simulated variance:", var( problem.2.sample ), "\n" )
```



\newpage
# Problem 3

### Part (a)

Suppose $W$ is an exponential random variable with an expected value of $W$ is $\hbox{E}[W] = 12.5$. What is the rate parameter $r$?

**Answer**

```{r}
cat( "Rate parameter:", 1/12.5, "\n" )
```



### Part (b)

Draw a random sample of 10,000 observations from an exponential distribution with the rate parameter that you determined in part (a) and store this result in a variable named `problem.3.sample`. Create a histogram of your data, and then superimpose the density curve for this exponential distribution.

**Answer**

```{r}
problem.3.sample <- rexp( 10000, rate = 1/12.5)

hist( problem.3.sample,
      main = "Problem 3 Exponential Histogram",
      xlab = "x",
      ylab = "Density",
      prob = TRUE,
      breaks = 50,
      col = "azure1")

curve( dexp(x, rate = 1/12.5), lwd = 3, col = "maroon", add = TRUE)
```

### Part (c)

Use your simulated data from part (b) to verify that the rate parameter you calculated in part (a) gives the correct expected value.

**Answer**

```{r}
mean( problem.3.sample )
```

\newpage
# Problem 4

In this problem, you must evaluate the integral:
$$
\int_{7.5}^{12.3} \frac{1}{10} e^{-x/10} dx\ =\ ???
$$

### Part (a)

What type of probability distribution is being integrated here? What are the parameters? What are the lower and upper limits of integration?

**Answer** This is an exponential distribution with rate parameter $r = 1/10$. The lower limit of integration is $x = 7.5$, and the upper limit of integration is $x = 12.3$. This integral represents the area between $x = 7.5$ and $x = 12.3$ beneath the density curve for an exponential distribution with rate parameter $r = 1/10$.




### Part (b)

Use an appropriate built-in probability function in R to evaluate this integral. (Hint: it starts with the letter 'p', and you'll have to use it twice with slightly different input values and perform a subtraction.)

**Answer**

```{r}
pexp( 12.3, rate = 1/10) - pexp(7.5, rate = 1/10)
```

### Part (c)

Construct a simulation to evaluate this integral.

**Answer**

```{r}
problem.4.sample <- rexp( 10000, rate = 1/10 )

mean( (problem.4.sample > 7.5) & (problem.4.sample <= 12.3) )
```


\newpage
# Problem 5

Now we're going to do this all over again, but for the gamma distribution. From lecture, we saw that a *gamma distribution* with shape parameter $a$ and rate parameter $r$ has the density function:
$$
f_X(x)\ =\ \frac{r^a \cdot x^{a-1} e^{-rx}}{\Gamma(a)},\ \ x \geq 0, a > 0, r > 0
$$
For this problem, let $X$ be a random variable that has a gamma distribution with shape parameter $a = 4$ and rate parameter $r = 1/9$.

### Part (a)

Draw a sample of 10,000 independent observations from this gamma distribution, and display a histogram of the values. Then superimpose a gamma density with a shape parameter $a = 4$ and rate parameter $r = 1/9$ by using the `curve()` function with the `add = TRUE` option.

**Answer**

```{r}
problem.5.sample <- rgamma(1000000, shape = 4, rate = 1/9 )

hist( problem.5.sample,
      main = "Histogram of Problem 5 Gamma Distribution",
      xlab = "x",
      ylab = "Density",
      prob = TRUE,
      breaks = 50,
      col = "azure2" )

curve( dgamma(x, shape = 4, rate = 1/9),
       lwd = 3,
       col = "darkslategray4",
       add = TRUE )
```

### Part (b)

Calculate the survival probability $S_X(62.7)\ =\ \Pr(X \geq 62.7)$ using the built-in R function `pgamma()`.

**Answer**

```{r}
pgamma(62.7, shape = 4, rate = 1/9, lower.tail = FALSE)
```


### Part (c)

Use your simulated data from part (a) to verify your calculation in part (b).

**Answer**

```{r}
mean( problem.5.sample > 62.5 )
```


### Part (d)

The expected value and variance of a gamma random variable with shape parameter $a$ and rate parameter $r$ are:
\begin{eqnarray*}
\hbox{E}[X] & = & \frac{a}{r}\\
\\
\hbox{Var}[X] & = & \frac{a}{r^2}
\end{eqnarray*}
Calculate the expected value and variance of a gamma random variable with shape parameter $a = 4$ and rate parameter $r = 1/9$.

**Answer**

```{r}
4/(1/9)

4/(1/9)^2
```



### Part (e)

Use your simulated data from part (a) to verify your calculations in part (d).

**Answer**

```{r}
mean( problem.5.sample )
var( problem.5.sample )
```



\newpage
# Problem 6

### Problem Statement

Let $Y$ be a continuous random variable with density function:
$$
f_Y(y)\ =\ \frac{(2.7)^{6.4} \cdot y^{5.4} e^{-2.7y}}{\Gamma(6.4)},\ \ y \geq 0
$$

### Part (a)

What is the name of the probability distribution of $X$? What are the parameters for this distribution?

**Answer** This is a gamma distribution with shape parameter $a = 6.4$ and rate parameter $r = 2.7$.




### Part (b)

Draw a random sample of 10,000 independent observations from the distribution $Y$, and call this sample `problem.6.sample'. Create a histogram for this distribution, and then superimpose the density curve for the distribution that you named in part (a).

**Answer**

```{r}
problem.6.sample <- rgamma( 10000, shape = 6.4, rate = 2.7)

hist( problem.6.sample,
      main = "Problem 6 Mystery Distribution Histogram",
      xlab = "y",
      ylab = "Density",
      prob = TRUE,
      breaks = 50,
      col = "salmon1")

curve( dgamma(x, shape = 6.4, rate = 2.7), lwd = 3, add = TRUE )
```

### Part (c)

Use the built-in R function `pgamma()` to calculate the cumulative probability at $x = 1.8$; that is, $F_Y(1.8)\ =\ \Pr(Y \leq 1.8)$.

**Answer**

```{r}
pgamma(1.8, shape = 6.4, rate = 2.7 )
```


### Part (d)

Use your simulated data from part (b) to verify your answer in part (c).

**Answer**

```{r}
mean( problem.6.sample <= 1.8 )
```


### Part (e)

Calculate the expected value and variance for the density in part (a).

**Answer**

```{r}
cat( "Expected value:", 6.4 / 2.7, "\n" )

cat( "Variance:", 6.4 / (2.7^2), "\n" )
```



### Part (e)

Use your simulated data from part (a) to verify your calculations of the expected value and variance of $X$ in part (d).

**Answer**

```{r}
cat( "Simulated expected value:", mean( problem.6.sample ), "\n" )

cat( "Simulated variance:", var( problem.6.sample ), "\n" )
```



\newpage
# Problem 7

Recall from Problem 5 that if $W$ is a random variable that has a gamma distribution with shape parameter $a$ and rate paramater $r$, then the expected value and variance are:
\begin{eqnarray*}
\hbox{E}[W] & = & \frac{a}{r}\\
\\
\hbox{Var}[W] & = & \frac{a}{r^2}
\end{eqnarray*}
Suppose you don't know either the shape parameter or the rate parameter, but you do know the actual numerical values of the expectation and variance of $W$:
\begin{eqnarray*}
\hbox{E}[X] & = & 5.7\\
\\
\hbox{Var}[X] & = & 12.2
\end{eqnarray*}

### Part (a)

Calculate the values of the shape parameter $a$ and the rate parameter $r$.

**Answer**

```{r}
problem.7.expectation <- 5.7
problem.7.variance <- 12.2

problem.7.shape.parameter <- problem.7.expectation^2 / problem.7.variance
problem.7.rate.parameter <- problem.7.shape.parameter / problem.7.expectation

cat( "Shape parameter:", problem.7.shape.parameter, "\n" )
cat( "Rate parameter:", problem.7.rate.parameter, "\n" )

```

### Part (b)

Show that the shape and rate parameters that you calculated in part (a) will give the correct values when you substitute them into the formulas for the expected value and the variance.

**Answer**

```{r}
cat( "Expected value:", problem.7.shape.parameter / problem.7.rate.parameter, "\n" )
cat( "Expected value:", problem.7.shape.parameter / problem.7.rate.parameter^2, "\n" )
```

### Part (c)

Draw a random sample of 10,000 independent observations from a gamma distribution with the shape and rate parameters that you calculated in part (a). Superimpose the density curve for the specified gamma distribution and then use this data to verify your calculations in part (b) by calculating the sample mean and sample variance.

**Answer**

```{r}
problem.7.sample <- rgamma( 1000000, 
                            shape = problem.7.shape.parameter,
                            rate = problem.7.rate.parameter )

hist( problem.7.sample,
      main = "Problem 7 Gamma Distribution",
      xlab = "x",
      ylab = "Density",
      breaks = 50,
      prob = TRUE,
      col = "navajowhite1")

curve( dgamma(x, shape = problem.7.shape.parameter, rate = problem.7.rate.parameter),
       lwd = 3,
       col = "navajowhite4",
       add = TRUE)
```

Now for the mean and variance:
```{r}
cat( "Expected value:", mean( problem.7.sample ), "\n" )
cat( "Variance:", var( problem.7.sample ), "\n" )
```







\newpage
# Problem 8

In this problem, we will evaluate the definite integral:
$$
\int_{2.3}^{7.1} \frac{(1.3)^{5.8} \cdot y^{4.8} e^{-1.3y}}{\Gamma(5.8)} \cdot dy\ =\ ???
$$

### Part (a)

Use an appropriate built-in probability function in R to evaluate this integral. (Hint: it starts with the letter 'p', and you'll have to use it twice with slightly different input values and then subtract.)

**Answer** This is an integral from $x = 2.3$ to $x = 7.1$ of the density function for a gamma distribution with shape parameter $a = 5.8$ and rate parameter $r = 1.3$.

```{r}
pgamma(7.1, shape = 5.8, rate = 1.3) - pgamma(2.3, shape = 5.8, rate = 1.3 )
```




### Part (b)

Construct a simulation that will approximately evaluate the integral, and compare your estimate with the value you calculated in part (a).

**Answer**

```{r}
problem.8.sample <- rgamma(1000000, shape = 5.8, rate = 1.3)

mean( (problem.8.sample > 2.3) & (problem.8.sample <= 7.1) )
```


### Part (c)

Using your simulated data from part (b), construct a histogram and then superimpose the density curve that you determined in part (a).

```{r}
hist( problem.8.sample,
      main = "Problem 8 Integral Evaluation",
      xlab = "x",
      ylab = "Density",
      prob = TRUE,
      breaks = 50,
      col = "lightsteelblue2")

curve( dgamma(x, shape = 5.8, rate = 1.3), lwd = 3,
       col = "lightsteelblue4", add = TRUE )
```







