---
title: "Problem Set 7"
subtitle: 'STAT E-80: Basic Probability Using R'
output:
  pdf_document: default
  html_document:
    df_print: paged
---


# Problem 1


A certain disease occurs in 1.5\% of the population. A diagnostic test for this disease has sensitivity 90\% and specificity 85\%.

###Part (a)


What is the probability of a false positive with this test?

**Answer**
```{r}
falsepositive<-1-.85
print(falsepositive)


```




###Part (b)

What is the probability of a false negative for this test?
**Answer**
1-.9 = .1

###Part (c)

Using Bayes' theorem, calculate the positive predictive value for this test.

**Answer**

```{r}
sensitivity <- 0.90
specificity <- 0.85
prevalence <- 0.015

answer <- (sensitivity * prevalence) / ( (sensitivity * prevalence) +
                                   ((1-specificity) * (1-prevalence) ) )
print(answer)

```

###Part (d)

Construct a simulation to verify that your answer in part (c) is correct. You are welcome to borrow the code from the Lecture 8 R notebook and modify it.

**Answer**
```{r}
number.replications <- 100000

# We specify the sensitivity, specificity, and prevalence just once,
# and then use these variables in the code. That way, if we want to
# use different values, it's easy to make the changes.

sensitivity <- 0.90
specificity <- 0.85
prevalence <- 0.015

# We need two result vectors, one for disease status,
# and the other for the test result.

disease.vector <- character( number.replications )
test.vector <- character( number.replications )

for( replication.index in 1:number.replications ) {
    
    # First, we simulate the disease status
    
    disease.status <- sample( c("Diseased", "Healthy"), 1,
                              prob = c(prevalence, 1-prevalence) )
    
    # Now we simulate the test result; this will depend on
    # the disease status of the patient
    
    if( disease.status == "Diseased" ) {
        
        # For a subject with disease, we use the sensitivity
        # to randomly sample the test result
        
        test.result <- sample( c("Positive", "Negative"), 1,
                               prob = c(sensitivity, 1 - sensitivity) )
        
    } else {
        
        # For a subject without disease, we use the specificity
        # to randomly sample the test result
        
        test.result <- sample( c("Positive", "Negative"), 1,
                               prob = c(1-specificity, specificity) )
    }
    
    # Now that we have the disease status and the test result,
    # let's store these in the result vectors:
    
    disease.vector[ replication.index ] <- disease.status
    test.vector[ replication.index ] <- test.result
}

# Finally, we use our standard 
filtered.vector <- disease.vector[ test.vector == "Positive" ]
mean( filtered.vector == "Diseased" )



```




# Problem 2

Justin uses hashtags in 70\% of his text messages, while Caitlin uses hashtags in 35\% of her messages. Caitlin sends 55\% of the text messages between the two, while Justin sends 45\% of the messages. Now you observe the message:

*yo wassup wuz u @ that crazee partee?!? lol*

### Part (a)

Using Bayes' theorem, calculate the conditional probability that Caitlin sent the message.

**Answer**
```{r}
prob.hashtag.justin <- 0.7
prob.hashtag.caitlin <- 0.35
prob.caitlin <- 0.55
prob.justin <- 1 - prob.caitlin

answer<-(prob.hashtag.caitlin * prob.caitlin) / 
    ( (prob.hashtag.caitlin * prob.caitlin) + (prob.hashtag.justin * prob.justin) )
print(answer)
```


###Part (b)

Construct a simulation to show that your calculation in part (a) is correct.

**Answer**



Now let's simulate this. This is essentially exactly the same as our simulation for the disease screening problem, except that we've changed the variable names to reflect the problem statement.
```{r}
number.replications <- 100000

prob.hashtag.justin <- 0.7
prob.hashtag.caitlin <- 0.35
prob.caitlin <- 0.55

# We need two result vectors, one to keep track of
# the sender of the message, and the other to keep track
# of the presence or absence of a hashtag.

sender.vector <- character( number.replications )
hashtag.vector <- character( number.replications )

# The sender of the message in this problem is analogous
# to the disease status in the screening problem, while
# the hashtag is analogous to the diagnostic test.

for( replication.index in 1:number.replications ) {
    
    # First, we randomly generate the sender of the message
    
    sender <- sample( c("Caitlin", "Justin"), 1,
                              prob = c(prob.caitlin, 1-prob.caitlin) )
    
    # Next, we randomly generate the presence or absence of a hashtag, given
    # that the sender
    
    if( sender == "Caitlin" ) {
        hashtag.status <- sample( c("Yes", "No"), 1,
                               prob = c(prob.hashtag.caitlin, 1 - prob.hashtag.caitlin) )
    } else {
        hashtag.status <- sample( c("Yes", "No"), 1,
                               prob = c(prob.hashtag.justin, 1-prob.hashtag.justin) )
    }
    
    # Now we store the sender and hashtag.status in result vectors:
    
    sender.vector[ replication.index ] <- sender
    hashtag.vector[ replication.index ] <- hashtag.status
    
}

# Finally, we use standard filtering method. First, we extract those
# elements of the sender vector that had a hashtag in their email.

filtered.vector <- sender.vector[ hashtag.vector == "Yes" ]

# Next, we determine the proportion of the filtered values
# that were really sent by Caitlin

mean( filtered.vector == "Caitlin" )
```



# Problem 3

An insurance company finds that 85\% of the policyholders are good drivers, and 15\% of the policyholders are bad drivers. A good driver has a 10\% probability of having a car accident in one year, while a bad driver has a 35\% probability of having a car accident in one year.

### Part (a)

The Chevalier de Mere had an accident last year. Use Bayes' theorem to calculate the probability that he is a bad driver.

**Answer**
```{r}


prob.gooddriver.wreck <- .1
prob.baddriver.wreck <- .35

prob.gooddriver <-.85
prob.baddriver <- 1-prob.gooddriver


answer<-(prob.baddriver.wreck * prob.baddriver) / 
    ( (prob.baddriver.wreck* prob.baddriver) + (prob.gooddriver.wreck * prob.gooddriver) )
print(answer)
```

```{r}
number.replications <- 100000

prob.gooddriver.wreck <- .1
prob.baddriver.wreck <- .35

prob.gooddriver <-.85
prob.baddriver <- 1-prob.gooddriver

# We need two result vectors, one to keep track of
# the driver of the message, and the other to keep track
# of the presence or absence of a driver.

driver.vector <- character( number.replications )
driver.status <- character( number.replications )
wreck.vector <- character( number.replications )


for( replication.index in 1:number.replications ) {
    
    # First, we randomly good or bad driver
    
    driver <- sample( c("BAD", "GOOD"), 1,
                              prob = c( 1-prob.gooddriver,prob.gooddriver) )
    
    # Next, we randomly generate the presence or absence of a driver, given
    # that the driver
    
    if( driver == "BAD" ) {
        driver.status <- sample( c("WRECKEDCAR", "DIDNTWRECKCAR"), 1,
                               prob = c(prob.baddriver.wreck, 1 - prob.baddriver.wreck) )
    } else {
        driver.status <- sample( c("WRECKEDCAR", "DIDNTWRECKCAR"), 1,
                               prob = c(prob.gooddriver.wreck, 1-prob.gooddriver.wreck) )
    }
    
    # Now we store the driver and driver.status in result vectors:
    
    driver.vector[ replication.index ] <- driver
    wreck.vector[ replication.index ] <- driver.status
    
}

# Finally, we use standard filtering method. First, we extract those
# elements of the driver vector that had a driver in their email.

filtered.vector <- driver.vector[ wreck.vector == "WRECKEDCAR" ]

# Next, we determine the proportion of the filtered values
# that were wrecked by bad drivers

mean( filtered.vector == "BAD" )
```





# Problem 4

In this problem, we will perform some simple explorations with the binomial distribution.

Suppose we have a binomial distribution with $n = 8$ trials and a probability of success of $p = 0.35$. 

### Part (a)

The standard probability formula for the binomial distribution is:
$$
\Pr(X = k)\ =\ {n \choose k} \cdot p^k \cdot (1-p)^{n-k}
$$
Using this formula, calculate the probability of observing exactly 4 successes for the binomial distribution specified above.
**Answer**
```{r}
answer=choose(8,4)*(.35)^4*(1-.35)^4
print(answer)
```






### Part (b)

Using the built-in function `dbinom()`, calculate the probability of observing exactly 4 successes for the binomial distribution specified above.

**Answer**
```{r}
dbinom( 4, 8, 0.35)
```



### Part (c)

Using the built-in function `pbinom()`, calculate the probability of observing no more than 3 successes for this binomial distribution.

**Answer**

```{r}
pbinom( 3, 8, 0.35)
```

### Part (d)

Calculate the expected value and variance of this binomial distribution.

**Answer**
```{r}
expectedvalue=.35*8
print (expectedvalue)
varience=8*.35*(1-.35)
print (varience)

```



### Part (e)

Construct a simulation to verify your answers for parts (a) through (d). In this simulation, you should directly model the concept that a binomial random variable is a sum of Bernoulli random variables. To do this, for each replication of the `for` loop you should use the `sample()` function to generate 8 independent Bernoulli random variables with probability of success $p = 0.35$, and then add them together. Do NOT use the `rbinom()` function to generate the random binomial value, because that's too easy -- instead, really try to capture the idea that the binomial distribution is just a sum of Bernoulli random variables.

**Answer**
```{r}
number.replications <- 10000

number.trials <- 8
probability.success <- 0.35

result.vector <- numeric( number.replications )

for( replication.index in 1:number.replications ) {
    
    random.value <- sum (random.sample <- sample( c(1,0), number.trials, replace = TRUE,
                             prob = c(probability.success, 1 - probability.success) ))
    
    result.vector[ replication.index ] <-  random.value }


cat ("calculate the probability of observing exactly 4 successes ")
cat (mean( result.vector==4 ))
cat ("calculate the probability of observing at least 3 successes ")
cat (mean( result.vector <=3 ))
cat( "Expected value (exact):", number.trials * probability.success, "\n" )
cat( "Expected value (simulated):", mean( result.vector ), "\n" )

cat( "Variance (exact):", number.trials * probability.success * (1-probability.success), "\n" )
cat( "Variance (simulated):", var( result.vector ), "\n" )


```



# Problem 5

Each day for 3 weeks, the philosopher Z'aph'ma goes to his local convenience store and purchases an 8-spot Massachusetts State Lottery Keno ticket.

### Part (a)

An 8-spot Keno ticket will win some money if at least 4 of the selected numbers match the 20-number sample. What is the probability that an individual 8-spot Keno ticket will win some money i.e.\ have a positive payout? Calculate this value using the `phyper()` function, and store the value in the variable `winning.keno.prob`. Remember that we're only interested in whether or not the ticket wins some money; we don't care about how much the payout is.

**Answer**
```{r}
winning.keno.prob <- 1 - pbinom(4, 20, 4/20)
print (winning.keno.prob)
```




### Part (b)

What is the probability that Z`aph`ma will have at least 3 winning tickets during the 3-week period that he purchases Keno tickets? Calculate this using the `pbinom()` function and the `winning.keno.prob` value that you calculated in part (a).

**Answer**


```{r}
1-pbinom(3,21,winning.keno.prob)

```



### Part (c)

Construct a simulation to verify your calculation in part (c):

* Start out the usual way, by defining the number of replications and a logical vector to store your simulation results. Then begin a `for` loop, and for each time through, do this:

    * To simulate playing Keno for 3 weeks, use the `rhyper()` function to draw 21 random values from a hypergeometric distribution. Store this random sample in a variable called `random.sample`.

    * Count how many elements in `random.sample` are greater than or equal to 4, and store this value in a variable called `number.winning.tickets`.

    * Test to see if `number.winning.tickets` is greater than or equal to 3, and if it is, store the value `TRUE` in your result vector.

* When the `for` loop is done, calculate the proportion of `TRUE` values in your result vector.


**Answer**


```{r}
number.replications <- 10000
winning.keno.prob <- 1 - pbinom(4, 20, 4/20)
print (winning.keno.prob)
number.trials <- 21
probability.success <- winning.keno.prob

result.vector <- logical( number.replications )

for(replication.index in 1:number.replications) {
    
    random.sample <- sample( c(1,0), number.trials, replace = TRUE,
                                       prob = c(probability.success, 
                                                1 - probability.success) )
    number.winning.tickets<-sum( random.sample )
    
    if( number.winning.tickets > 3 ) {
        result.vector[replication.index ] <- TRUE
    }
}

mean( result.vector )
answer <- mean( result.vector )
cat ("Z'aph'ma will get 3 or more winning tickets with a simulated probablity of " , answer  )
```


# Problem 6

In the game of Chuck-A-Luck, the player places a \$1 wager, selects a number from 1 to 6 and then rolls 3 dice. If the selected number comes up exactly once out of the three dice, the player gets the initial wager back, along with an additional \$1. If the selected number comes up exactly twice out the three dice, then the player gets the initial wager back, along with an additional \$2. Finally, if the selected number comes up on all three of the dice, the player gets the initial wager back, plus an additional \$10.

###Part (a)

Calculate the expected profit for Chuck-A-Luck. Use the `dbinom()` function to calculate the selected number coming up once, twice, or three times on the three dice.

**Answer**
```{r}
profit.vector=c(-1,1,2,10)
probablity.vector<-
                  c(dbinom( 0, 3, (1/6)),
                  dbinom( 1, 3, (1/6)),
                  dbinom( 2, 3, (1/6)),
                  dbinom( 3, 3, (1/6)))



expected.profit <- sum(probablity.vector*profit.vector)
cat ("expected profit = " ,expected.profit)
```


###Part (b)

Construct a simulation to verify that your calculation in part (a) of the expected profit was correct. Again, do not use the `rbinom()` function, but instead use `sample()` to directly model the action of rolling three dice.

**Answer**
```{r}
# First, we'll set the number of replications of the game:

number.replications <- 10000

# Next, we create the vector that will hold the results of each experiment. Remember
# that when we create a vector with the logical() function, all the elements have
# the initial value of FALSE.

result.vector <- numeric( number.replications )

# Now we can run a for loop over all replications

for( replication.index in 1:number.replications ) {
  

  chuck.a.luck.value <- sum(sample(c(1,0),3,replace = TRUE,prob=c(1/6,5/6)))
  if (chuck.a.luck.value==0){
     result.vector[ replication.index ] <- -1
  }
  else if (chuck.a.luck.value==1){
     result.vector[ replication.index ] <- 1
  }
  else if (chuck.a.luck.value==2){
     result.vector[ replication.index ] <- 2
  }
  else if (chuck.a.luck.value==3){
     result.vector[ replication.index ] <- 10
  }
  
  
}

mean( result.vector )
```








# Problem 7

We usually specify a binomial distribution by assigning values to $n$, the number of trials, and $p$, the probability of success. Suppose you don't know either $n$ or $p$, but you do know that the expected value of the distribution is 6, and also that the variance is 1.5:
\begin{eqnarray*}
\hbox{E}[X] & = & 6\\
\\
\hbox{Var}[X] & = & 1.5
\end{eqnarray*}

### Part (a)

Using the values of the expected value and variance, calculate the values of $n$ and $p$. Hint: look in the lecture for the algebraic expressions for the mean and variance of a binomial distribution, and then think about what the algebraic expression is for the variance divided by the mean.

**Answer**

first I always wondered what something like this was suppose to mean
\begin{eqnarray*}
\hbox{E}[X] & = & 6\\
\\
\hbox{Var}[X] & = & 1.5
\end{eqnarray*}

what is hbox suppose to mean ?


lets call that v and execped value = E
v=n * P * ( 1 - P )
E=N*P
For variance
solve for P
v=n * P * ( 1 - P )
1.5=n * P * ( 1 - P )
-10np^2+10np-15=0
I can then solve this with quadradic formula,  but I would get two values,  so this one just doesn't make much sense, can't wait to see solution

Expected value

Solve for N
np(1/p)=v
np=v(1-p)
n=(v(1-p))/p
Expected value
much easier
p=E/N
N=E/p










### Part (b)

Using the values of $n$ and $p$ that you obtained in part (a), conduct a simulation to show that the resulting binomial distribution really does have an expected value of 6 and a variance of 1.5. Once again, don't use the built-in `rbinom()` function to generate random binomial values, but instead use the `sample()` function to generate $n$ independent Bernoulli random variables, each with common probability of success $p$. Once you've finished the simulation, calculate the mean and variance of the `result.vector` and show that they are very close to the values specified in the problem statement.

**Answer**
I don't see how this works given a quadratic solution for part a,  will wait for soltuion





# Problem 8

Chicken Curry takes 40 shots, and for each shot he scores a basket with probability $p = 0.42$. Kobe Beef is a total ball hog, and he takes 55 shots, but for each each shot he scores a basket with probability $p = 0.35$. Assume that all the shots are independent of one another. What is the probability that Kobe will score more baskets than Chicken? Answer this by constructing a simulation. You'll have to determine how to model Chicken's number of baskets, as well as Kobe's number of baskets, and then what to do with the results.

**Answer**
```{r}
# First, we'll set the number of replications of the game:

number.replications <- 10000



result.vector.kobescoremore<-numeric ( number.replications )
result.vector.tie<- numeric( number.replications )
# Now we can run a for loop over all replications

for( replication.index in 1:number.replications ) {
  
 
  kobe.score<- sum(sample(c(1,0),40,replace = TRUE,prob=c(.58,.42)))
  chicken.score<- sum(sample(c(1,0),55,replace = TRUE,prob=c(.35,.65)))
  kobe.total<-sum(kobe.score)
  chicken.total<-sum(chicken.score)

  if (kobe.total>chicken.total){
  result.vector.kobescoremore[ replication.index ]<-1

    result.vector.tie[ replication.index ]<-0
  }
  else if (kobe.total<chicken.total){
  result.vector.kobescoremore[ replication.index ]<-0
  result.vector.tie[ replication.index ]<-0
  }
  else {result.vector.tie[ replication.index ]<-1}
  
  
}

kobe.scoresmore.prob<-mean( result.vector.kobescoremore)
tie<-mean( result.vector.tie)
cat ("probablity of tie = ")
cat (tie)
cat ("\n")
cat("Kobe will score more with probablity" ,kobe.scoresmore.prob)
cat ("\n")
cat("Chicken will score more with probablity" ,1-kobe.scoresmore.prob-tie)
```





























